---
title: "Exploratory Data Analysis"
subtitle: "Credit Risk Intelligence Platform"
author: "Nihira Sharma"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(DBI)
library(lubridate)
library(knitr)
```

------------------------------------------------------------------------

## 1. Connect to Postgres & Load Data

```{r connect}
library(DBI)
library(RPostgres)

# Connect directly
con <- dbConnect(
  RPostgres::Postgres(),
  user = "credit_app",
  password = Sys.getenv("CREDIT_DB_PASSWORD"),
  host = "localhost",
  port = 5432,
  dbname = "credit_risk"
)

# Test connection
dbGetQuery(con, "SELECT version();")
```

------------------------------------------------------------------------

## 2. Load All 7 Tables

```{r load_data}
# Load main application table
app <- dbGetQuery(con, "SELECT * FROM application_train") %>% as_tibble()

# Load child tables
prev_app <- dbGetQuery(con, "SELECT * FROM previous_application") %>% as_tibble()
bureau <- dbGetQuery(con, "SELECT * FROM bureau") %>% as_tibble()
bureau_bal <- dbGetQuery(con, "SELECT * FROM bureau_balance") %>% as_tibble()
credit_card <- dbGetQuery(con, "SELECT * FROM credit_card_balance") %>% as_tibble()
installments <- dbGetQuery(con, "SELECT * FROM installments_payments") %>% as_tibble()
pos_cash <- dbGetQuery(con, "SELECT * FROM pos_cash_balance") %>% as_tibble()

cat("All 7 tables loaded successfully\n")
```

------------------------------------------------------------------------

## 3. High-Level Overview

### 3.1 Row Counts & Table Dimensions

```{r overview}
overview <- tibble(
  Table = c("application_train", "previous_application", "bureau", "bureau_balance",
            "credit_card_balance", "installments_payments", "pos_cash_balance"),
  Rows = c(nrow(app), nrow(prev_app), nrow(bureau), nrow(bureau_bal),
           nrow(credit_card), nrow(installments), nrow(pos_cash)),
  Columns = c(ncol(app), ncol(prev_app), ncol(bureau), ncol(bureau_bal),
              ncol(credit_card), ncol(installments), ncol(pos_cash)),
  Grain = c("One row per applicant (SK_ID_CURR)",
            "One row per previous application (SK_ID_PREV)",
            "One row per bureau credit record (SK_ID_BUREAU)",
            "One row per bureau-month (SK_ID_BUREAU, MONTHS_BALANCE)",
            "One row per card-month (SK_ID_PREV, SK_ID_CURR, MONTHS_BALANCE)",
            "One row per installment payment",
            "One row per POS-month (SK_ID_PREV, SK_ID_CURR, MONTHS_BALANCE)")
)

kable(overview, caption = "Table Overview & Grain")
```

------------------------------------------------------------------------

## 4. Target Variable Analysis

### 4.1 Default Rate

```{r target_dist}
target_summary <- app %>%
  summarise(
    Total_Applicants = n(),
    Defaults = sum(target, na.rm = TRUE),
    Non_Defaults = sum(1 - target, na.rm = TRUE),
    Default_Rate = mean(target, na.rm = TRUE) * 100
  )

kable(target_summary, caption = "Target Distribution: Default Rate")

# Visualize
ggplot(app, aes(x = factor(target))) +
  geom_bar(fill = c("#2ecc71", "#e74c3c"), alpha = 0.8) +
  scale_x_discrete(labels = c("Non-Default", "Default")) +
  labs(title = "Target Distribution",
       subtitle = "Class Balance in application_train",
       x = "Target", y = "Count") +
  theme_minimal() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)
```

**Key insight:** \~8% default rate - highly imbalanced, typical for credit data.

------------------------------------------------------------------------

## 5. Missingness Analysis

### 5.1 Missing Data in application_train

```{r missingness_app, fig.height=12, fig.width=9}
missing_app <- app %>%
  summarise(across(everything(), 
                   list(missing_pct = ~sum(is.na(.)) / n() * 100))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_pct") %>%
  mutate(column = str_remove(column, "_missing_pct")) %>%
  filter(missing_pct > 0) %>%
  arrange(desc(missing_pct))

kable(missing_app, caption = "application_train: Columns with Missing Values")

# Visualize high-missing columns (>40%)
missing_app %>%
  filter(missing_pct > 40) %>%
  ggplot(aes(x = reorder(column, -missing_pct), y = missing_pct)) +
  geom_col(fill = "#e67e22", alpha = 0.8) +
  labs(title = "High-Missing Columns in application_train",
       subtitle = ">40% missing may need special handling",
       x = "Column", y = "% Missing") +
  coord_flip() +
  theme_minimal()
```

------------------------------------------------------------------------

## 6. Numeric Features - Univariate Distributions

### 6.1 Income, Credit Amount, Annuity

```{r numeric_explore}
numeric_cols <- c("amt_income_total", "amt_credit", "amt_annuity", 
                  "days_birth", "days_employed", "cnt_children")

numeric_summary <- app %>%
  select(all_of(numeric_cols)) %>%
  summarise(across(everything(),
                   list(
                     Min = ~min(., na.rm = TRUE),
                     Median = ~median(., na.rm = TRUE),
                     Mean = ~mean(., na.rm = TRUE),
                     Max = ~max(., na.rm = TRUE),
                     SD = ~sd(., na.rm = TRUE)
                   )))

# Transpose for readability
numeric_summary_t <- numeric_summary %>%
  pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("column", "metric"), sep = "_") %>%
  pivot_wider(names_from = "metric", values_from = "value")

kable(numeric_summary_t, caption = "Numeric Features: Summary Statistics", digits = 2)
```

### 6.2 Key Derived Features

```{r derived}
app <- app %>%
  mutate(
    age_years = -days_birth / 365.25,
    employment_years = case_when(
      days_employed == 365243 ~ NA_real_,  # Home Credit's "unemployed" sentinel
      TRUE ~ -days_employed / 365.25
    ),
    income_per_family = amt_income_total / (cnt_fam_members + 1),
    credit_to_income = amt_credit / amt_income_total,
    annuity_to_income = amt_annuity / amt_income_total,
    credit_to_goods = amt_credit / amt_goods_price
  )

# Summary of derived features
derived_summary <- app %>%
  select(age_years, employment_years, credit_to_income, annuity_to_income) %>%
  summarise(across(everything(),
                   list(Mean = ~mean(., na.rm = TRUE),
                        Median = ~median(., na.rm = TRUE),
                        SD = ~sd(., na.rm = TRUE))))

kable(derived_summary, caption = "Derived Features: Summary", digits = 2)
```

#### Age Distribution

```{r age_plot}
ggplot(app, aes(x = age_years)) +
  geom_histogram(binwidth = 5, fill = "#3498db", alpha = 0.8) +
  labs(title = "Age Distribution",
       subtitle = "Derived from DAYS_BIRTH",
       x = "Age (years)", y = "Count") +
  theme_minimal()
```

#### Employment Years (post-sentinel fix)

```{r employment_plot}
ggplot(app %>% filter(!is.na(employment_years)), aes(x = employment_years)) +
  geom_histogram(binwidth = 2, fill = "#9b59b6", alpha = 0.8) +
  labs(title = "Employment Duration Distribution",
       subtitle = "Excludes 365243 sentinel (unemployed)",
       x = "Employment Years", y = "Count") +
  theme_minimal()
```

------------------------------------------------------------------------

## 7. Child Table Grains & Aggregation Potential

### 7.1 Bureau Records per Applicant

```{r bureau_grain}
bureau_per_app <- bureau %>%
  group_by(sk_id_curr) %>%
  summarise(n_bureau_records = n(), .groups = 'drop')

bureau_grain_summary <- tibble(
  Metric = c("Applicants with bureau history", "Mean bureau records/applicant",
             "Median", "Max", "Min"),
  Value = c(nrow(bureau_per_app), 
            mean(bureau_per_app$n_bureau_records),
            median(bureau_per_app$n_bureau_records),
            max(bureau_per_app$n_bureau_records),
            min(bureau_per_app$n_bureau_records))
)

kable(bureau_grain_summary, caption = "Bureau Records per Applicant", digits = 2)
```

### 7.2 Previous Applications per Applicant

```{r prev_app_grain}
prev_app_per_app <- prev_app %>%
  group_by(sk_id_curr) %>%
  summarise(n_prev_apps = n(), .groups = 'drop')

prev_app_grain_summary <- tibble(
  Metric = c("Applicants with prior apps", "Mean prior apps/applicant",
             "Median", "Max"),
  Value = c(nrow(prev_app_per_app),
            mean(prev_app_per_app$n_prev_apps),
            median(prev_app_per_app$n_prev_apps),
            max(prev_app_per_app$n_prev_apps))
)

kable(prev_app_grain_summary, caption = "Previous Applications per Applicant", digits = 2)
```

### 7.3 Installments per Previous Application

```{r installments_grain}
inst_per_prev <- installments %>%
  group_by(sk_id_prev) %>%
  summarise(n_installments = n(), .groups = 'drop')

inst_grain_summary <- tibble(
  Metric = c("Previous apps with installments", "Mean installments/prev_app",
             "Median", "Max"),
  Value = c(nrow(inst_per_prev),
            mean(inst_per_prev$n_installments),
            median(inst_per_prev$n_installments),
            max(inst_per_prev$n_installments))
)

kable(inst_grain_summary, caption = "Installment Records per Previous Application", digits = 2)
```

------------------------------------------------------------------------

## 8. Early Target Relationships (Visual Only)

### 8.1 Default Rate by Income Band

```{r target_by_income}
app %>%
  mutate(income_band = cut(amt_income_total, 
                           breaks = quantile(amt_income_total, 
                                            probs = seq(0, 1, 0.25), 
                                            na.rm = TRUE),
                           include.lowest = TRUE)) %>%
  group_by(income_band) %>%
  summarise(default_rate = mean(target, na.rm = TRUE) * 100, n = n(), .groups = 'drop') %>%
  ggplot(aes(x = income_band, y = default_rate)) +
  geom_col(fill = "#e74c3c", alpha = 0.8) +
  geom_text(aes(label = paste0(round(default_rate, 1), "%")), vjust = -0.5) +
  labs(title = "Default Rate by Income Quartile",
       x = "Income Band (Quartiles)", y = "Default Rate (%)") +
  theme_minimal()
```

### 8.2 Default Rate by Age Band

```{r target_by_age}
app %>%
  mutate(age_band = cut(age_years, breaks = seq(20, 80, 10))) %>%
  group_by(age_band) %>%
  summarise(default_rate = mean(target, na.rm = TRUE) * 100, n = n(), .groups = 'drop') %>%
  ggplot(aes(x = age_band, y = default_rate)) +
  geom_col(fill = "#3498db", alpha = 0.8) +
  geom_text(aes(label = paste0(round(default_rate, 1), "%")), vjust = -0.5) +
  labs(title = "Default Rate by Age Band",
       x = "Age Band (years)", y = "Default Rate (%)") +
  theme_minimal()
```

------------------------------------------------------------------------

## 9. Key Findings & Data Quality Notes

```{r findings}
findings <- tibble(
  Finding = c(
    "Default rate: ~8%",
    "DAYS_EMPLOYED = 365243 sentinel",
    "High-missing columns: EXT_SOURCE_*, building features",
    "No absolute dates: uses anonymized day-offsets",
    "Child tables: no orphaned records detected",
    "Installments table: highest volume (~13.6M rows)"
  ),
  Impact = c(
    "Class imbalance: use stratified splits, not SMOTE",
    "Recode as NA in feature engineering",
    "Plan imputation strategy in Phase 2",
    "Cannot do true out-of-time validation",
    "Data integrity: safe to proceed",
    "Will need efficient SQL aggregation"
  )
)

kable(findings, caption = "Key EDA Findings")
```

------------------------------------------------------------------------

## 10. Close Connection

```{r close}
dbDisconnect(con)
cat("Connection closed\n")
```

------------------------------------------------------------------------
