---
title: "Scorecard Generation"
subtitle: "Credit Risk Intelligence Platform"
author: "Nihira Sharma"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    highlight: pygments
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=TRUE, results='markup'}
knitr::opts_chunk$set(echo=TRUE, message=TRUE, warning=FALSE,
                      fig.width=10, fig.height=6, results='markup')
library(tidyverse)
library(scales)
library(knitr)
source("../R/model_utils.R")   # metric functions
source("../R/scorecard.R")     # scorecard functions

cat("Scorecard Generation - Ready\n")
cat("Loaded model_utils.R and scorecard.R\n")
```

---

## Overview

This notebook converts the production logistic regression model into an **industry-standard bin-level points scorecard** - the format lenders actually deploy.

- **Objective:** Translate each WOE bin into a fixed number of points, score every applicant, set Approve/Review/Decline cutoffs, and confirm the scorecard reproduces the model.
- **Inputs:** `outputs/model_logistic.rds`, `data/processed/test_with_predictions.rds`, `data/processed/woe_bins.rds`
- **Scaling:** PDO method (Points to Double the Odds)
- **Output:** A points card where each feature/bin is worth whole points, plus a scaled score (~300–850) where **higher = safer**

The value of a bin-level card is auditability: a credit analyst can read a single applicant's decision row by row - "age band gives +28, employment band gives +12, …" - and the points sum to the score. This is the artifact real risk teams sign off on.

---

# Section 1: Load Model, Predictions & WOE Bins

```{r load_artifacts, results='markup'}
model_lr  <- readRDS("../outputs/model_logistic.rds")
test_pred <- readRDS("../data/processed/test_with_predictions.rds")
woe_bins  <- readRDS("../data/processed/woe_bins.rds")

coefs <- coef(model_lr)

cat("Model loaded\n")
cat("  Coefficients:", length(coefs), "(intercept + features)\n")
cat("  Test set:", nrow(test_pred), "applicants\n")
cat("  WOE bins object class:", class(woe_bins)[1], "\n\n")

stopifnot("pred_lr" %in% names(test_pred), "target" %in% names(test_pred))

cat("Predicted probability range:",
    round(min(test_pred$pred_lr), 4), "–",
    round(max(test_pred$pred_lr), 4), "\n")
```

---

# Section 2: PDO Scaling Parameters

A scorecard maps log-odds to points with a linear transform:

$$\text{Score} = \text{Offset} + \text{Factor} \times \ln(\text{odds}_\text{good})$$

anchored by three business choices - a **base score**, the **base odds** at that score, and **PDO** (points to double the odds). `Factor` and `Offset` are derived from these.

```{r pdo_params, results='markup'}
scaling <- make_scaling(base_score = 600, base_odds = 50, pdo = 20)

cat("PDO Scaling Parameters\n")
cat("──────────────────────────────────────\n")
cat("  Base score:     ", scaling$base_score, "\n")
cat("  Base odds:      ", scaling$base_odds, ": 1 (good:bad)\n")
cat("  PDO:            ", scaling$pdo, "points to double odds\n\n")
cat("  Derived Factor: ", round(scaling$factor, 4), "\n")
cat("  Derived Offset: ", round(scaling$offset, 4), "\n\n")
cat("Interpretation: every", scaling$pdo, "points, the odds of being a\n")
cat("good (non-default) applicant double.\n")

saveRDS(scaling, "../outputs/scorecard_scaling_params.rds")
```

---

# Section 3: Build the Bin-Level Points Card

For each feature and bin:

$$\text{Points} = \frac{\text{Offset} - \text{Factor} \times \beta_0}{n} - \text{Factor} \times \beta_i \times \text{WOE}_\text{bin}$$

The intercept is spread evenly across features so an applicant's matched bin points sum to their total score.

```{r build_card, results='markup'}
card <- build_scorecard(coefs, woe_bins, scaling)

base_pts <- attr(card, "base_points_per_feature")

cat("Scorecard built\n")
cat("  Rows (feature × bin):", nrow(card), "\n")
cat("  Features in card:    ", length(unique(card$coef_name)), "\n")
cat("  Base points/feature: ", round(base_pts, 1), "\n\n")

cat("Sample of the points card:\n")
print(head(card %>% select(variable, bin, woe, beta, points), 20))

write_csv(card, "../outputs/scorecard_points_card.csv")
cat("\nSaved outputs/scorecard_points_card.csv\n")
```

## Points Range by Feature

How much each feature can swing an applicant's score - the spread between its lowest- and highest-points bins. Wider bars are the most influential features.

```{r feature_swing, results='markup'}
feature_swing <- card %>%
  group_by(variable) %>%
  summarise(
    min_points = min(points),
    max_points = max(points),
    swing = max(points) - min(points),
    .groups = "drop"
  ) %>%
  arrange(desc(swing)) %>%
  slice_head(n = 15)

ggplot(feature_swing, aes(x = reorder(variable, swing), y = swing)) +
  geom_col(fill = "#C9A227", alpha = 0.9) +
  geom_text(aes(label = swing), hjust = -0.2, size = 3.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Score Influence by Feature",
    subtitle = "Points swing between each feature's weakest and strongest bin",
    x = NULL, y = "Points Swing (max − min bin)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Section 4: Score the Test Population

Because the test set already holds WOE values, scoring is a direct, exact transform of the model logit (guaranteed identical to summing card points).

```{r score_test, results='markup'}
test_scored <- test_pred %>%
  mutate(score = score_from_woe(., coefs, scaling))

cat("Score Distribution (Test Set)\n")
cat("──────────────────────────────────────\n")
cat("  Min:   ", min(test_scored$score), "\n")
cat("  Q1:    ", round(quantile(test_scored$score, 0.25)), "\n")
cat("  Median:", round(median(test_scored$score)), "\n")
cat("  Mean:  ", round(mean(test_scored$score)), "\n")
cat("  Q3:    ", round(quantile(test_scored$score, 0.75)), "\n")
cat("  Max:   ", max(test_scored$score), "\n\n")

cor_check <- cor(test_scored$score, test_scored$pred_lr)
cat("Correlation(score, P(default)):", round(cor_check, 4), "\n")
cat("(Should be strongly negative - higher score = lower risk)\n")
```

## Score Distribution

```{r score_hist, results='markup'}
ggplot(test_scored, aes(x = score)) +
  geom_histogram(binwidth = 10, fill = "#C9A227", colour = "white", alpha = 0.9) +
  geom_vline(aes(xintercept = median(score)), linetype = "dashed", colour = "#2E2E2E") +
  labs(
    title = "Credit Score Distribution",
    subtitle = "Test set - higher scores indicate lower default risk",
    x = "Credit Score", y = "Number of Applicants"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## Score vs. Observed Default Rate

The key validation chart: as scores rise, the actual default rate must fall monotonically.

```{r score_vs_default, results='markup'}
score_bands_check <- test_scored %>%
  mutate(score_band = cut(score,
                          breaks = seq(floor(min(score)/50)*50,
                                       ceiling(max(score)/50)*50, by = 50),
                          include.lowest = TRUE)) %>%
  group_by(score_band) %>%
  summarise(n = n(), default_rate = mean(target), .groups = "drop") %>%
  filter(n >= 30)

ggplot(score_bands_check, aes(x = score_band, y = default_rate, fill = default_rate)) +
  geom_col(alpha = 0.9) +
  geom_text(aes(label = percent(default_rate, accuracy = 0.1)),
            vjust = -0.5, size = 3.2, colour = "#2E2E2E") +
  scale_fill_gradient(low = "#2E6B3E", high = "#8B2E2E", guide = "none") +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Observed Default Rate by Score Band",
    subtitle = "Monotonic decline confirms the scorecard ranks risk correctly",
    x = "Score Band", y = "Observed Default Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# Section 5: Decision Cutoffs (Approve / Review / Decline)

Cutoffs convert the continuous score into an actionable decision. Adjust `decline_cut` / `approve_cut` after seeing your score distribution above so the approved band lands at a default rate you accept.

```{r decision_cutoffs, results='markup'}
# Cutoffs are calibrated to THIS model's score range (≈500–621)
# Scores compress because Gini ≈ 0.31; we split on the actual distribution,
# not a generic 300–850 assumption. Tune these to your risk appetite
decline_cut <- 545   # below this: DECLINE
approve_cut <- 575   # at/above this: APPROVE; between: REVIEW

test_decisions <- test_scored %>%
  mutate(decision = assign_decision(score, decline_cut, approve_cut))

decision_summary <- test_decisions %>%
  group_by(decision) %>%
  summarise(
    n_applicants = n(),
    pct_population = n() / nrow(test_decisions),
    n_defaults = sum(target),
    default_rate = mean(target),
    .groups = "drop"
  ) %>%
  mutate(
    pct_population = round(pct_population, 4),
    default_rate = round(default_rate, 4),
    pct_of_all_defaults = round(n_defaults / sum(n_defaults), 4)
  )

cat("Decision Band Summary\n")
cat("──────────────────────────────────────\n")
cat("  Decline: score <", decline_cut, "\n")
cat("  Approve: score >=", approve_cut, "\n")
cat("  Review: ", decline_cut, "–", approve_cut - 1, "\n\n")
print(decision_summary)

approve_band <- decision_summary %>% filter(decision == "APPROVE")
decline_band <- decision_summary %>% filter(decision == "DECLINE")

cat("\nBusiness Read-out:\n")
cat("  Auto-approve rate:    ", percent(approve_band$pct_population, 0.1), "\n")
cat("  Approved default rate:", percent(approve_band$default_rate, 0.01),
    "(vs", percent(mean(test_decisions$target), 0.01), "overall)\n")
cat("  Declined band catches:", percent(decline_band$pct_of_all_defaults, 0.1),
    "of all defaults\n")

write_csv(decision_summary, "../outputs/scorecard_decision_summary.csv")
cat("\nSaved outputs/scorecard_decision_summary.csv\n")
```

## Decision Bands Visualised

```{r decision_plot, results='markup'}
ggplot(test_decisions, aes(x = score, fill = decision)) +
  geom_histogram(binwidth = 10, colour = "white", alpha = 0.9) +
  geom_vline(xintercept = c(decline_cut, approve_cut),
             linetype = "dashed", colour = "#2E2E2E") +
  scale_fill_manual(values = c("DECLINE" = "#8B2E2E",
                               "REVIEW"  = "#C9A227",
                               "APPROVE" = "#2E6B3E")) +
  labs(
    title = "Lending Decisions by Credit Score",
    subtitle = "Decline / Review / Approve bands on the test population",
    x = "Credit Score", y = "Number of Applicants", fill = "Decision"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Section 6: Score Band Reference Table

The day-to-day lookup a risk team uses: each band with population share, default rate, and recommended action.

```{r score_band_table, results='markup'}
score_band_table <- test_decisions %>%
  mutate(band = cut(score,
                    breaks = c(-Inf, 520, 540, 555, 570, 585, 600, Inf),
                    labels = c("<520", "520–539", "540–554", "555–569",
                               "570–584", "585–599", "600+"))) %>%
  group_by(band) %>%
  summarise(n = n(),
            pct_population = n() / nrow(test_decisions),
            default_rate = mean(target),
            .groups = "drop") %>%
  mutate(
    cumulative_pct = cumsum(n) / sum(n),
    recommended_action = case_when(
      default_rate >= 0.12 ~ "DECLINE",
      default_rate >= 0.05 ~ "REVIEW",
      TRUE                 ~ "APPROVE"
    ),
    pct_population = percent(pct_population, 0.1),
    default_rate = percent(default_rate, 0.01),
    cumulative_pct = percent(cumulative_pct, 0.1)
  )

cat("Score Band Reference Table\n\n")
print(score_band_table)

write_csv(score_band_table, "../outputs/scorecard_band_reference.csv")
cat("\nSaved outputs/scorecard_band_reference.csv\n")
```

---

# Section 7: Validate the Scorecard Preserves Model Performance

A scorecard is a monotonic rescaling of the model's probabilities, so its ranking power (Gini/KS/AUC) must match the model exactly. Any discrepancy signals a scaling bug.

```{r validate_scorecard, results='markup'}
# Higher score = lower risk, so rank on NEGATIVE score for the metric helpers.
gini_score <- compute_gini(test_scored$target, -test_scored$score)
ks_score   <- compute_ks(test_scored$target, -test_scored$score)
auc_score  <- compute_auc(test_scored$target, -test_scored$score)

gini_model <- compute_gini(test_scored$target, test_scored$pred_lr)
ks_model   <- compute_ks(test_scored$target, test_scored$pred_lr)
auc_model  <- compute_auc(test_scored$target, test_scored$pred_lr)

validation <- tibble(
  Metric = c("Gini", "KS", "AUC"),
  `From Model (prob)` = round(c(gini_model, ks_model, auc_model), 4),
  `From Scorecard`    = round(c(gini_score, ks_score, auc_score), 4),
  Difference          = round(c(gini_score - gini_model,
                                ks_score - ks_model,
                                auc_score - auc_model), 6)
)

cat("Scorecard vs. Model - Ranking Power\n")
cat("──────────────────────────────────────\n")
print(validation)

if (all(abs(validation$Difference) < 0.001)) {
  cat("\nScorecard preserves model performance (differences ≈ 0)\n")
  cat("  The points scaling is a faithful monotonic transform.\n")
} else {
  cat("\nScorecard metrics differ from model - check scaling.\n")
}

write_csv(validation, "../outputs/scorecard_validation.csv")
```

---

# Section 8: Save Scorecard Artifacts

```{r save_artifacts, results='markup'}
test_scored_out <- test_decisions %>%
  select(SK_ID_CURR, target, pred_lr, score, decision)

saveRDS(test_scored_out, "../data/processed/test_scored.rds")
write_csv(test_scored_out, "../scorecard_output/test_scored.csv")

# Self-contained, deployable scorecard object for the dashboard
scorecard_obj <- list(
  scaling        = scaling,
  card           = card,
  cutoffs        = list(decline = decline_cut, approve = approve_cut),
  band_reference = score_band_table,
  validation     = validation,
  model_coefs    = coefs
)
saveRDS(scorecard_obj, "../outputs/scorecard.rds")

cat("\nScorecard artifacts saved\n")
cat("  - outputs/scorecard.rds (deployable scorecard object)\n")
cat("  - outputs/scorecard_points_card.csv (the bin-level card)\n")
cat("  - outputs/scorecard_scaling_params.rds\n")
cat("  - outputs/scorecard_decision_summary.csv\n")
cat("  - outputs/scorecard_band_reference.csv\n")
cat("  - outputs/scorecard_validation.csv\n")
cat("  - data/processed/test_scored.rds\n")
cat("  - scorecard_output/test_scored.csv\n")
```

---

# Summary

This notebook converted the logistic regression model into a deployable **bin-level points scorecard**:

- **PDO scaling** (base 600, 50:1 odds, 20 PDO) maps log-odds to an intuitive score where higher means safer.
- **The points card** assigns each WOE bin a whole-number point value - every decision is auditable row by row.
- **Decision cutoffs** (Decline < 545, Review 545–574, Approve ≥ 575) translate scores into lending actions with quantified default rates per band.
- **Validation** confirms the scorecard reproduces the model's exact ranking power (Gini/KS/AUC).

The scorecard object (`outputs/scorecard.rds`) is ready for the interactive dashboard, where a user can score a hypothetical applicant and see the points breakdown.

---